---
title: "Regional Flood Skew by SPDE using Bayesian INLA"
author: "Dave Holtschlag"
date: " `r format(Sys.Date(), '%A %B %d, %Y')` "
output: html_document
bibliography: ../text/bibliography.bib
---


```{r setup, include=FALSE}
library(tidyverse)
library(ggthemes)
library(INLA)
library(rgeos)
library(maptools)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

A preliminary Bayesian spatial model of flood skew values in hydrologic unit regions 04 (U.S. Great Lakes) and parts of 05 (Ohio River) are developed to illustrate a spatial modeling approach within the R statistical environment using Integrated Nested Laplace Approximation (R-INLA) (@JSSv063i19). 

@JSSv063i19

## Approach

Let $s$ be any location in study area with a random spatial effect $U(s)$ at that location.  $U(s)$ is a  stochastic process with $s \in \mathbf{D}, \space where \space \mathbf{D} \in \Re^2$. A set of $n$ realizations of $U(s)$ could be written as $u(s_i), i=1,2,...,n$. It is commonly assumed that $u(s)$ has a multivariate Gaussian distribution.  If $U(s)$ is continuous over $\mathbf{D}$, we have a continuously indexed Gaussian field $(GF)$.  Such a $GF$ is fully specified by its mean and variance. 

Instead of directly observing the $GF$ at $s$, are measurements generally include measurement error such that:
$$ y(s_i) = u(s_i) + e_i $$
An isotopic spatial correlation function on a stationary random field can be developed to describe the average decrease in correlation with increasing Euclidean separation distance. Commonly, the range of separation distance $h$ among sites is discretized into a uniform set of intervals $\delta h$ that collectively span the range. Sample correlations are then computed for each interval, and a monotonically decreasing function of the midpoint of each separation interval and the corresponding sample correlations.  

The $Mat\acute{e}rn$ family of correlatiohn functions is seeing increasing use in spatial statistical modeling (@guttorp_miscellanea_2006). The $Mat\acute{e}rn$ correlation function $Cor_M$ has a scale parameter $\kappa > 0$, and a smoothness parameter $\nu > 0$. For two locations $\mathbf{s}_i$ and $\mathbf{s}_j$, the stationary and isotropic $Mat\acute{e}rn$ correlation function is:
$$Cor_M(U(\mathbf{s}_i), U(\mathbf{s}_j))) = \frac{2^{i-\nu}}{\Gamma(\nu)}(\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)^\nu K_\nu (\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)$$

## Read and plot flood skew statistics

To illustrate the development of a 

```{r read_skew_data}
df    <- read_tsv('../data/skew_0405.txt', comment = '#') 

# plot density of station skew

df %>% 
   ggplot( aes( x = sta_skew)) +
   geom_histogram() +
   theme_few() +
   labs( x = 'Station flood skew', y = 'Count', 
         title = 'Distribution of station flood skew in hydrologic regions 04 and 05')

# Eliminate skews greater than 2
ndx_drop <- which(df$sta_skew > 2)
df       <- df[-ndx_drop,]
```



```{r plot_map, fig.width = 10}
upper_midwest <- map_data('state', region = c('ohio', 'michigan', 'indiana','illinois', 
                                              'wisconsin', 'pennsylvania', 'new york',
                                              'kentucky','west virginia'))

bound_box <- data.frame(longitude   = c(-87.1,-77.6, -77.6, -87.1, -87.1),
                       latitude    = c( 37.5, 37.5,  45.8,  45.8,  37.5),
                       region = rep(1, 5))

upper_midwest %>%
   ggplot( aes(x = long, y = lat, group = region)) +
   geom_polygon(color = 'gray', fill = NA) +
   theme_few() +
   coord_map(projection = 'albers', lat1 = 37.5, lat2 = 45.8) +
   geom_point(data  = df,
              aes(x = longitude, y = latitude, group = NULL, fill = sta_skew), 
              size = 2.0, shape = 21, color = 'black', stroke = 0.2) +
   scale_fill_gradient2() +
   labs(x = 'Longitude', y = 'Latitude') +
   geom_polygon(data = bound_box,
                aes(x = longitude, y = latitude, group = NULL), 
                color = 'blue', fill = NA, linetype = 'dashed') +
   theme(legend.position = 'bottom')
   

```

## Limit points to bounding box


```{r subset_box, fig.width = 10, fig.height = 9}
# ID points outside of bounding box
ndx_drop <- which( df$longitude > -77.6 | df$longitude < -87.1 | df$latitude > 45.8 )

df_area <- df[-ndx_drop, ]

# which <- map_data('state', region = 'michigan' && subregion = 'south')

mi    <- map_data('state', region = 'michigan')
ndx_s <- which(mi$subregion == 'south')
lp    <- mi[ndx_s,]

upper_mid  <- map_data('state', region = c('ohio', 'indiana', 
                                              'kentucky','west virginia'))

study_area <- rbind(lp, upper_mid)

bound_box <- data.frame(longitude = c(-87.1, -80.0, -78.6, -77.6, -77.6, -81.5, -85.0, -87.1, -87.1),
                        latitude  = c( 37.5,  37.5,  39.0,  40.5,  43.5,  44.8,  45.8,  42.0,  37.5),
                        region    = rep(1, 9))

study_area %>%
   ggplot( aes(x = long, y = lat, group = region)) +
   geom_polygon(color = 'gray', fill = NA) +
   theme_few() +
   # coord_map(projection = 'albers', lat1 = 37.5, lat2 = 45.8) +
   geom_point(data  = df_area,
              aes(x = longitude, y = latitude, group = NULL, fill = sta_skew), 
              size = 2.0, shape = 21, color = 'black', stroke = 0.2) +
   scale_fill_gradient2() +
   labs(x = 'Longitude', y = 'Latitude') +
   geom_polygon(data = bound_box,
                aes(x = longitude, y = latitude, group = NULL), 
                color = 'blue', fill = NA, linetype = 'dashed') +
   scale_x_continuous( limits = c(-87.5, -77.5)) +
   scale_y_continuous( limits = c( 37.5,  46.0)) +
   theme(legend.position = 'bottom')

```

## Finite element mesh assuming convex hull

```{r gen_mesh, fig.height = 9, fig.width = 9}

# femesh <- inla.mesh.2d(, bound_box[,1:2], max.edge = c(0.5, 1.0), n = 4, offset = c(0.3, 0.3))

femesh <- inla.mesh.2d(, bound_box[,1:2], max.edge = c(0.4, 0.5), offset = c(0.2, 0.5))

plot(femesh, asp = 1)
points(df_area[,c('longitude', 'latitude')], col = 'red', pch = 16)
points(bound_box[,1:2], type = 'l', col = 'blue')

```

## Estimation of a Response Surface

The initial estimator includes an intercept term with parameter $\beta_0$ and a spatial effect $\mathbf{A}u$ as:

$$ \eta^* = \mathbf{1} \beta_0 + \mathbf{A}u $$
where $u$ $\thicksim$ $GF(0,\Sigma)$, where $GF$ is a Gaussian field.  

```{r }

```


# References

```{r generateBibliography, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
require("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
read.bibtex(file = "../text/bibliography.bib")
```

