---
title: "Bayesian Modeling of Spatial and Spatial-Temporal Processes by use of the SPDE Approach"
author: "Dave Holtschlag"
date: " `r format(Sys.Date(), '%A %B %d, %Y')` "
output: html_document
bibliography: ../text/bibliography.bib
---


```{r setup, include=FALSE}
library(tidyverse)
library(ggthemes)
library(INLA)
library(rgeos)
library(maptools)
library(MASS)
library(kableExtra)
library(gridExtra)
library(lattice)
options(width = 132)
knitr::opts_chunk$set(echo = TRUE)
```

## Purpose and Scope

The purpose of this analysis is to provide a simple, preliminary example of the SPDE (Stochastic Partial Differential Equation) approach for estimating a spatially varying hydrologic processes by use of the Bayesian Integrated Nested Laplace Approximation (INLA) [@rue2017bayesian] algorithm in the R statistical programming environment. The presentation is intended to simulate further interest in applications related to processes generating annual flood series, understanding how regulation and urbanization may be impacting the annual flood series, and the attribution of trends in annual peak flows. For example, a spatial-temporal analysis of annual peaks could be developed from long-term unregulated and non-urbanized basins. Estimates from this model could be compared with peak flow statistics from regulated and urbanized areas to assess expected affects. The logit transform of the annual exceedance probabilities may be a suitable metric for spatial-temporal analysis of annual peak flows. Covariates, such as antecedent rainfall, could be included in a SPDE analysis to potentially increase model sensitivity.  


## Backround and Introduction  

Bayesian analysis can be developed based on Markov chain Monte Carlo (MCMC) simulation or numerical approximation.  INLA is a numerical approximation algorithm that is computationally efficient and a much faster alternative to MCMC that scales well to large data sets ^[Morrison, Kathryn, 2017, A gentle INLA tutorial', Precision Analytics, Montreal, Quebec, Canada, December 20, https://www.precision-analytics.ca/blog-1/inla, accessed March 27, 2018)], yet provide comparable precision. In particular, MCMC simulations for a network containing 1,000 streamgages with an average of 40 annual peaks may not be computationally feasible. In addition, the relative simplicity and adaptability of the R-INLA interface facilitates application to complex spatial-temporal models [@JSSv063i19].   

For reasons of convenience only, a readily available and related hydrologic data set of station skews is used to illustrate the general SPDE approach.  Station skews represent the coefficient of skewness in the distribution of annual peak flows computed by use of the Expected Moments Algorithm [@b17c]. In this analysis, a subset of streamgages in hydrologic unit regions 04 (U.S. Great Lakes) and parts of 05 (Ohio River) was used to illustrate the application. 


## Approach

In this preliminary analysis, the response variable $y(s)$ is the skew of the distribution of the $log_{10}$ transformed annual peaks at a streamgage monitoring flow from a basin with centroid at $s$. Here, $s$ is spatially indexed by latitude and longitude. The spatial variation in skew is partitioned into a mean (intercept) component $beta_0$, the linear projection, with projector matrix $\mathbf{A}$, of the estimated Gaussian random field at $u(s)$, and a measurement error $e$, as indicated below. 

(@linearproj) 
$$ y(s) = \mathbf{1} \beta_0 + \mathbf{A} u(s) + e $$ 

To include covariates, the first term on the right hand side of equation [1], $\mathbf{1} \beta_0$, would be replaced with $\mathbf{F^T} \cdot \mathbf{\beta}$, where $\mathbf{F}$ is a matrix of covariates, $\beta$ is a parameter vector.  

The covariance of the marginal distribution of $y(s)$, is
(@covar_y)
$$\Sigma_y = \Sigma + \sigma_e^2 \mathbf{I}$$
where $\Sigma$ is the joint covariance for all $u(\mathbf{s_i})$ $i = 1,2,...n$ a realization from $U(s)$ at $n$ locations. A continuous stochastic process $U(s)$ indexed over space is Gaussian random field $(GRF)$.  Samples $u(s)$ from a $GRF$ are generally assumed to have a multivariate Gaussian distribution. Thus, the distribution of $u(s)$ can be fully specified by its mean and covariance.    
Each ${i,j}$ element of $\Sigma$ is 
(@Sig_ij)
$$\Sigma_{i,j} = \sigma_u^2 Cor_M(U(s_i), U(s_j)))$$
where $\sigma_u^2$ is the marginal variance of $u(s)$, and $Cor_M(U(s_i), U(s_j))$ is the $Matern$ correlation function, which can be written as:    

(@Cor_M)
$$Cor_M(U(\mathbf{s}_i), U(\mathbf{s}_j))) = \frac{2^{1-\nu}}{\Gamma(\nu)}(\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)^\nu K_\nu (\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)$$
where $\parallel \cdot \parallel$ indicates the Euclidean distance, $K_{\nu}$ is the modified Bessel function of the second kind, $\kappa$ is a scale and $\nu$ is a smoothness parameter of the $Cor_M$.  The $Cor_M$ was used in place of correlation defined on the basis of monotonically decreasing function through paired samples at increasing separation distance $h$.   

For convenience, the error component will be considered indepenent and identically distributed as $N(0,\sigma^2)$. It should be noted that the reciprocal of variance $\frac{1}{\sigma^2}$ is referred to as the precision, and is the form of uncertainty estimated in this analysis.   More generally, covariates, such as basin characteristics or climatic variables, could be added to the estimation process. 

be any location in study area with a random spatial effect $U(s)$ at that location.  $U(s)$ is a  stochastic process with $s \in \mathbf{D}, \space where \space \mathbf{D} \in \Re^2$. A set of $n$ realizations of $U(s)$ could be written as $u(s_i), i=1,2,...,n$. It is commonly assumed that $u(s)$ has a multivariate Gaussian distribution.  If $U(s)$ is continuous over $\mathbf{D}$, we have a continuously indexed Gaussian field $(GF)$.  Such a $GF$ is fully specified by its mean and variance. 


Assuming that the measurement errors at streamgages are independent for all $i \neq j$ and $e_i \sim N(0, \sigma _e^2)$ for all $i$. 

An isotopic spatial correlation function on a stationary random field can be developed to describe the average decrease in correlation with increasing Euclidean separation distance. Commonly, the range of separation distance $h$ among sites is discretized into a uniform set of intervals $\delta h$ that collectively span the range. Sample correlations are then computed for each interval, and a monotonically decreasing function of the midpoint of each separation interval and the corresponding sample correlations.  

The $Mat\acute{e}rn$ family of correlation functions is seeing increasing use in spatial statistical modeling (@guttorp_miscellanea_2006). The $Mat\acute{e}rn$ correlation function $Cor_M$ has a scale parameter $\kappa > 0$, and a smoothness parameter $\nu > 0$. For two locations $\mathbf{s}_i$ and $\mathbf{s}_j$, the stationary and isotropic $Mat\acute{e}rn$ correlation function is:
$$Cor_M(U(\mathbf{s}_i), U(\mathbf{s}_j))) = \frac{2^{i-\nu}}{\Gamma(\nu)}(\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)^\nu K_\nu (\kappa \parallel \mathbf{s}_i - \mathbf{s}_j \parallel)$$
## Applying the SPDE approach to streamgage data

### Read and plot streamgages in U.S. Great Lakes and Ohio River Basins with skew statistics

Annual peak flows are systematically obtained at USGS streamgages.  Statistical analyses of the annual peak flow series provides estimates of the mean, standard deviation, and skewness coefficient (skew) for the $log_{10}$ transformed flows. The skew coefficient indicates the asymmetry of the distribution of annual peaks.  Estimates of annual exceedance probabilities (AEP) for large magnitude floods are highly sensitive to skew estimates. The distribution of station skews for selected streamgages indicates a generally symmetric density with two apparent outliers with values greater than 2.  These two observations were deleted from the data set. After their removal, the distribution of skews appears consistent with a normal distribution. 

The plot show distribution of sample points $\mathbf s$ in the HUC 04 and 05 areas.  A model boundary (in dashed blue) is shown on the study area to restrict interpolation to areas with sufficient data.


```{r plot_study_area, fig.width = 10, fig.height = 9, message = FALSE, warning = FALSE, fig.cap = 'Figure 1. Study area in the Great Lakes - Ohio River basin showing model boundary as dashed blue polygon and flood skew statistics at selected streamgages as shaded circles'}

# Read location and skew data 
df    <- read_tsv('../data/skew_0405.txt', comment = '#') 



upper_midwest <- map_data('state', region = c('ohio', 'michigan', 'indiana','illinois', 
                                              'wisconsin', 'pennsylvania', 'new york',
                                              'kentucky','west virginia'))

bound_box <- data.frame(longitude = c(-87.1, -80.0, -78.6, -77.6, -77.6, -81.5, -85.0, -87.1, -87.1),
                        latitude  = c( 37.5,  37.5,  39.0,  40.5,  43.5,  44.8,  45.8,  42.0,  37.5),
                        region    = rep(1, 9))

upper_midwest %>%
   ggplot( aes(x = long, y = lat, group = region)) +
   geom_polygon(color = 'gray', fill = NA) +
   theme_few() +
   coord_map(projection = 'albers', lat1 = 37.5, lat2 = 45.8) +
   geom_point(data  = df,
              aes(x = longitude, y = latitude, group = NULL, fill = sta_skew), 
              size = 2.0, shape = 21, color = 'black', stroke = 0.2) +
   scale_fill_gradient2() +
   labs(x = 'Longitude', y = 'Latitude') +
   geom_polygon(data = bound_box,
                aes(x = longitude, y = latitude, group = NULL), 
                color = 'blue', fill = NA, linetype = 'dashed') +
   theme(legend.position = 'bottom')

```


### Identifying streamgages with unusual skew values

In figure 1, note two streamgages in northern Ohio that have large positive skew values

```{r read_skew_data, message = FALSE, fig.cap = 'Figure 2. Modelled area in the Great Lakes - Ohio River Area showing variation in flood skew statistics at selected streamgages'}

# plot density of station skew

df %>% 
   ggplot( aes( x = sta_skew)) +
   geom_histogram() +
   theme_few() +
   geom_vline( xintercept = 2, color = 'red', linetype = 'dashed') +
   annotate('text', x = 3.2, y = 5, label = 'Outliers', color = 'red') +
   labs( x = 'Coefficient of skewness for annual peak flow data', y = 'Count', 
         title = 'Distribution of flood skew coefficients in hydrologic regions 04 and 05')
```

### Remove streamgages identified as outliers

A break in the linearity between theoretical and sample quantiles is noted at theoretical quantiles below 0.9.

```{r drop_outliers, fig.cap = 'Figure 3. Distribution of skew values  for selected streamgages in the modelled area of the U.S. Great Lakes and Ohio River with outliers removed'}

# Eliminate skews greater than 2
ndx_drop <- which(df$sta_skew > 2)
df1      <- df[-ndx_drop,]

# ID points outside of model area
ndx_drop <- which( df1$longitude > -77.6 | df1$longitude < -87.1 | df1$latitude > 45.8 )
df_area <- df1[-ndx_drop, ]

# Compute normal distribution for mean and variance of selected sites
params   <- fitdistr(df_area$sta_skew, 'normal')$estimate

df_area %>% 
   ggplot( aes(sample = sta_skew)) +
   stat_qq( distribution = qnorm, dparams = params, color = 'blue') +
   theme_few() +
   labs(y = 'Station Skews', x = 'Theoretical Normal Quantiles',
        title = 'Normal Quantile - Sample Quantile Plot of Station Flood Skew',
        subplot = 'A break in the linearity between theoretical and sample quantiles is noted below 0.9') 

```

### Constrain study area by model boundary 

```{r subset_box, fig.width = 10, fig.height = 9, fig.cap = 'Figure 4. Modelled area in the Great Lakes - Ohio River Area showing variation in flood skew statistics at selected streamgages'}

# which <- map_data('state', region = 'michigan' && subregion = 'south')

mi    <- map_data('state', region = 'michigan')
ndx_s <- which(mi$subregion == 'south')
lp    <- mi[ndx_s,]

upper_mid  <- map_data('state', region = c('ohio', 'indiana', 
                                              'kentucky','west virginia'))

study_area <- rbind(lp, upper_mid)


study_area %>%
   ggplot( aes(x = long, y = lat, group = region)) +
   geom_polygon(color = 'gray', fill = NA) +
   theme_few() +
   # coord_map(projection = 'albers', lat1 = 37.5, lat2 = 45.8) +
   geom_point(data  = df_area,
              aes(x = longitude, y = latitude, group = NULL, fill = sta_skew), 
              size = 3.0, shape = 21, color = 'black', stroke = 0.2) +
   scale_fill_gradient2() +
   labs(x = 'Longitude', y = 'Latitude') +
   geom_polygon(data = bound_box,
                aes(x = longitude, y = latitude, group = NULL), 
                color = 'blue', fill = NA, linetype = 'dashed') +
   scale_x_continuous( limits = c(-87.5, -77.5)) +
   scale_y_continuous( limits = c( 37.5,  46.0)) +
   theme(legend.position = 'bottom')

```

### Finite element mesh on a convex area

Sample data are generally not located on a regular grid, but are distributed irregularly through the model area.  A finite element mesh (FEM) is used to span the model area with linear triangles, and is used to solve the SPDE. The FEM is composed of an inner domain containing all the sample points, and an outer extension.  The extension is needed to avoid a boundary effect that doubles the variance at the border relative to the  inner domain.  
This solution is highly sensitive to the parameters of the FEM, including the $max.edge$ and $offset$. The $max.edge$ parameter vector specifies the maximum allowed triangle edge lengths in the inner domain and the outer extension, respectively, in the same scale as the coordinates. The $offset$ parameter vector 


```{r gen_mesh1, fig.height = 9, fig.width = 9}

# Coordinates of the data points
coords    <- as.matrix(df_area[, c('longitude', 'latitude')])

femesh <- inla.mesh.2d(coords, loc.domain = bound_box[,1:2], max.edge = c(0.4, 0.5), 
                       offset = c(0.2, 0.5), cutoff = 0.05, min.angle = 22)

plot(femesh, asp = 1)
text(-77.3, 45.2,'Streamgage')
lines(c(-78.47414, -77.5), c(43.24588, 45.0), col = 'red', lwd = 2)
points(df_area[,c('longitude', 'latitude')], col = 'red', pch = 16)

text(-77.7, 37.7, 'Model boundary')
lines(c(-78.6, -77.5), c(39.0, 38.0), col = 'blue', lwd = 2)
lines(bound_box[,1:2], col = 'blue', lty = 2)

text(-77.5, 46.0, 'Inner domain')
lines(c(-79.5, -77.5), c(44.4, 45.8), col = 'forestgreen', lwd = 2)

text(-87.3, 46.0, 'Outer extension')
lines(c(-86.5, -87.5), c(44.9, 45.8), col = 'forestgreen', lwd = 2)




```


## Estimation of a Response Surface

The initial estimator includes an intercept term with parameter $\beta_0$ and a spatial effect $\mathbf{A}u$ as:

$$ \eta^* = \mathbf{1} \beta_0 + \mathbf{A}u $$
where $u$ $\thicksim$ $GF(0,\Sigma)$, where $GF$ is a Gaussian field.  

```{r SPDE_model_id}

# Use boundary file
# bound_box is the set of points forming a polygon defining the domain of the estimation area
# Use previously generated finite element mesh
# femesh <- inla.mesh.2d(, bound_box[,1:2], max.edge = c(0.4, 0.5), offset = c(0.2, 0.5))

# the SPDE model in the original parameterization is build using the function inla.spde2.matern()

spde1 <- inla.spde2.pcmatern(
   mesh   = femesh,   # mesh
   # alpha = 2 is the default for the inla.spde2.matern() function: 0 < alpha <= 2
   # related to smoothness parameter of the process. See Nov2017_tutorial, p. 16
   alpha   = 2,    
   # Hyperparameters
   prior.range = c(2.5, 0.50), # P(practic.range < 0.3 ) = 0.5
   prior.sigma = c(1.0, 0.01)  # P(sigma > 1) = 0.1
)

```


```{r Projector_matrix}
# The projector matrix A1 will project the random field modeled at the mesh nodes.  
#  
A1        <- inla.spde.make.A( femesh , loc = coords )

```

The inla.stack() function is useful for organizing data, covariates, indices, and projector matrices.  It also helps control the way effects are projected in the linear predictor.  The inla.stack function automatically eliminates any column in a projector matrix that has a zero sum, and generates a new simplified matrix. Note, the inla.stack.A() function extract a simplified predictor matrix to use with the inla() function, while the inla.stack.data() function extract the corresponding data. 

```{r data_stack, eval = TRUE}

stk1 <- inla.stack( data = list( resp    = df_area$sta_skew), 
                                 A       = list( A1, 1),
                                 effects = list(i = 1:spde1$n.spde,
                                                m = rep(1, nrow(df_area))),
                                 tag     = 'est')

```

Execution of dim(inla.stack.A( stk1 )) returns `r dim(inla.stack.A( stk1 ))` as compared to the dimension of A1 `r dim(A1)`.

Estimation of model parameters

```{r model_fit, eval = TRUE}
res1   <- inla(resp ~ 0 + m + f(i, model = spde1), 
               data = inla.stack.data(stk1), 
               control.predictor = list(A = inla.stack.A(stk1)))

```

Estimate of the fixed component $\beta_0$, is `r format(res1$summary.fix[1], digits = 4)`, which compares closely with the mean of station skews in the model area of `r format(mean(df_area$sta_skew), digits = 4)`.  The precision of the Gaussian observations, $\frac{1}{\sigma_e^2}$, was estimated as the first hyperparameter below as `r format(res1$summary.hyperpar[[1]][1], digits = 4)`.


```{r model_results, eval = TRUE, out.width = 10}
# Estimate of B0
print(res1$summary.fix)

# Estimate of the hyperparameters
print(res1$summary.hyperpar)
```

### SPDE Estimates at Streamgages



```{r spde_estimates}

# Estimates at streamgages from first model
df_area$pred_skew_spde <- drop(A1 %*% res1$summary.random$i$mean)

df_area[1:10, c('sta_no', 'rec_length', 'sta_skew', 'pred_skew_spde')] %>% 
   knitr::kable(digits = 4, format = 'html', table.attr = "style = 'width:60%;'") #  %>% 
   # kable_styling(full_width = F)

# Square Pearson correlation coefficient
r2 <- pearson <- format( cor(df_area$sta_skew, df_area$pred_skew_spde)^2, digits = 4)

df_area %>% 
   ggplot( aes( x = sta_skew, y = pred_skew_spde)) +
   geom_point() +
   theme_few()  +
   labs( x     = 'Computed station skew',
         y     = 'SPDE estimated skew',
         title = 'Relation between computed and estimated skew in HUC 04 and 05') +
   geom_abline( intercept = 0, slope = 1, color = 'blue') +
   scale_x_continuous( breaks = seq(-1.5, 1.5, 0.50)) +
   scale_y_continuous( breaks = seq(-0.5, 0.5, 0.25)) +
   annotate('text', x = -1.30, y = 0.5, 
            label = as.character(expression(paste(r^2,' = '))), parse = TRUE) +
   annotate('text', x = -1.05, y = 0.5, label = r2)

```

### SPDE Projection on the Study Area

```{r proj_grid}

long_lim <- c(-87.1, -77.5)
lat_lim  <- c( 37.5,  45.6)

xlim     <- seq(long_lim[1], long_lim[2], by = 0.1); 
ylim     <- seq(lat_lim[1],  lat_lim[2],  by = 0.1)

loc_mat <- as.matrix(expand.grid(x = xlim, y = ylim))
colnames(loc_mat) <- c('Longitude', 'Latitude')

nrow_mat <- length(xlim)
ncol_mat <- length(ylim)

pgrid0 <- inla.mesh.projector(femesh, xlim = long_lim, ylim = lat_lim,
                            dims = c( nrow_mat, ncol_mat ))

prd0.m <- inla.mesh.project(pgrid0, res1$summary.random$i$mean)
prd0.s <- inla.mesh.project(pgrid0, res1$summary.random$i$sd)

stkgrid  <- inla.stack(data    = list( resp = NA ), A = list( pgrid0$proj$A, 1),
                       effects = list( i   = 1:spde1$n.spde,
                                       m   = rep(1, nrow_mat * ncol_mat)),
                                       tag = 'prd.gr')

stk.all  <- inla.stack( stk1, stkgrid)

# Rebuild model for grid
res1g    <- inla(resp ~ 0 + m + f(i, model = spde1),
                 data = inla.stack.data( stk.all ), 
                 control.predictor = list(A = inla.stack.A( stk.all ),
                                          compute = TRUE ), quantiles = NULL,
                 control.results = list(return.marginals.random    = FALSE,
                                        return.marginals.predictor = FALSE))
# Print the CPU
print(res1g$cpu)

# Get indices for grid
igr  <- inla.stack.index( stk.all, 'prd.gr')$data
```

### SPDE Regional Skew Estimate in HUC 04 and HUC 05 

```{r plot_grid, fig.width = 10, fig.height = 9}

levelplot( matrix( res1g$summary.fitted[igr,1], nrow_mat),
           xlab = 'Matrix column (Longitude)', ylab = 'Matix row (Latitude)',
           main = 'SPDE expected skew mean',
           col.regions = topo.colors(99), scales = list(draw = TRUE))

levelplot( matrix(res1g$summary.fitted[igr, 2], nrow_mat),
           xlab = 'Matrix Column', ylab = 'Matrix Row', 
           main = 'SPDE expected skew standard deviation',
                        col.regions = topo.colors(99), scales = list(draw = TRUE))

# mean of skew in grid
# Overall mean of grid
spde_mean   <- mean(mean(matrix( res1g$summary.fitted[igr,1], nrow_mat)))
# De-meaned spde estimates of skew 
spde_demean <- matrix( res1g$summary.fitted[igr,1], nrow_mat) - spde_mean
# Studentized spde estimates of skew ( (x-mu)/sd )
spde_stdnt  <- spde_demean / matrix(res1g$summary.fitted[igr, 2], nrow_mat)

# Relate columns and rows to longitude and latitude
long2col <- function(long){
   x <-  10 * long + 775 + 96.5
   return(x)
}

lat2row  <- function(lat){
   y <-  10 *  lat + -374.5
   return(y)
}

myplot <- contourplot( matrix(spde_stdnt, nrow_mat),
           xlab = 'Matrix Column', ylab = 'Matrix Row', 
           main = 'SPDE Studentized Mean Skew',
                        col.regions = topo.colors(99), scales = list(draw = TRUE))


update(myplot, panel = function(...) {
   panel.contourplot(...)
   panel.xyplot(long2col(coords[,1]), lat2row(coords[,2]), pch = 16, col = "red")
})


```


## Summary and Conclusions

A Stochastic P

## References

```{r generateBibliography, echo=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
require("knitcitations")
cleanbib()
options("citation_format" = "pandoc")
read.bibtex(file = "../text/bibliography.bib")
```

